Appendix for: Deep Reinforcement Learning for Conservation Decisions

### 2021-04-29 manuscript/appendix.Rmd
:pen:

scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
pal <- ggthemes::solarized_pal()(8)
txtcolor <- "#586e75"
```

@@

At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.

@@

## Python dependencies loaded via R
sb3         <- import ("stable_baselines3")
gym         <- import ("gym")
gym_fishing <- import("gym_fishing")
gym_conservation <- import("gym_conservation")
```

@@

```{r}
## reproducible settings
np <- import("numpy")
seed <- 24L # integer
np$random$seed(seed)
# Optionally set to "" to force CPU-evaluation if needing perfect reproducibility
Sys.setenv("CUDA_VISIBLE_DEVICES"="")
set.seed(seed)
```

@@

```{r }
## initialize the environment
env <- gym$make("fishing-v1", sigma = 0.1)
```

@@

```{r}
a2c <- sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
```

@@

```{r}
# Simulate management under the trained agent
a2c <- sb3$A2C$load("cache/a2c")
```

@@

## represent the initial state size in the 'rescaled' state space.
state <- env$get_state( 0.75 )
state
```

@@

  out <- a2c$predict(state)
  action <- out[[1]]
  result <- env$step(action)
  state <- result[[1]]

@@

```{r}
a2c_sims <- env$simulate(a2c, reps = 500L)
```

@@

```{r}
a2c_policy <- env$policyfn(a2c, reps = 50L)
```

@@

```{r}
# Simulate management under the trained agent
td3 <- sb3$TD3$load("cache/td3")
td3_sims <- env$simulate(td3, reps = 500L)
td3_policy <- env$policyfn(td3, reps = 50L)
```

@@

```{r}
# Simulate under the optimal solution (given the model)
opt <- gym_fishing$models$escapement(env)
opt_sims <- env$simulate(opt, reps = 500L)
opt_policy <- env$policyfn(opt)
```

@@

```{r}
sims_df <- bind_rows(td3_sims, a2c_sims, opt_sims, .id = "model") %>%
  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])

policy_df <- bind_rows(td3_policy, a2c_policy, opt_policy, .id = "model") %>%
  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])

gamma <- 1 #discount
reward_df <- sims_df %>% 
  group_by(rep, model) %>%
  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
  group_by(time, model) %>%
  summarise(mean_reward = mean(cum_reward), 
            sd = sd(cum_reward), .groups = "drop")

@@

```{r}
ymin <- function(x) last(x[(ntile(x, 20)==1)])
ymax <- function(x) last(x[(ntile(x, 20)==19)])

fig_sims <- 
sims_df %>% 
  group_by(time, model) %>% 
  summarise(ymin = ymin(state),
            ymax = ymax(state),
            state = mean(state), .groups = "drop") %>%
  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
```

```{r}
fig_policy <- 
  policy_df %>% ggplot(aes(state, action, 
                           group=interaction(rep, model),
                           col = model)) + 
  geom_line(show.legend = FALSE) + 
  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) 
```

@@

fig_reward <- reward_df %>% 
  ggplot(aes(time, mean_reward)) + 
  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
                  ymax = mean_reward + 2*sd, fill = model),
              alpha=0.25, show.legend = FALSE) + 
  geom_line(aes(col = model), show.legend = FALSE) + 
  ylab("reward")
```

@@

```{r}
# Simulate management under the trained agent
env <- gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
td3 <- sb3$TD3$load("cache/td3")

@@

```{r}
hake <- read_csv("../data/hake.csv")
x0 <- hake %>%
 filter(year == min(year)) %>%
  pull(scaled_biomass)
Tmax <- 15
years <-1986:2000

td3_state <- td3_action <- numeric(Tmax)
td3_state[1] <- x0
td3_action[1] <- NA

@@

```{r}
env$reset()
## represent the initial state size in the 'rescaled' state space.
state <- env$get_state( x0 )

@@

  # RL-recommended harvest action:
  out <- td3$predict(state)
  action <- out[[1]]

  # Record state and resulting action
  td3_state[i] <- env$get_fish_population(state)
  td3_action[i] <- env$get_quota(action)

  # Implement RL-recommended harvest. 
  result <- env$step(action)
  state <- result[[1]]

  #state <- env$get_state(hake$scaled_biomass[i+1])
}

df <- bind_rows(
  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))

@@

```{r message=FALSE, fig.cap="Bifurcation diagram for tipping point scenario. The ecosystem begins in the desirable 'high' state under an evironmental parameter (e.g. global mean temperature, arbitrary units) of 0.19.  In the absence of conservation action, the environment worsens (e.g. rising mean temperature) as the parameter increases.  This results in only a slow degredation of the stable state, until the parameter crosses the tipping point threshold at about 0.215, where the upper stable branch is anihilated in a fold bifurcation and the system rapidly transitions to lower stable branch, around state of 0.1.   Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
bifur_df <- read_csv("figs/bifur.csv", col_types = "dccd")
bifur_df %>%
  ggplot(aes(parameter, state, lty=equilibrium, group = group)) + 
  geom_line()

@@

```{r}
env <- gym$make("conservation-v6")
```

@@

OU <- sb3$common$noise$OrnsteinUhlenbeckActionNoise
action_noise = OU(mean = np$zeros(1L),  sigma = noise_std * np$ones(1L))

@@

model <- sb3$TD3$load("cache/td3-conservation")
TD3_sims <- env$simulate(model, reps = 100L)
TD3_policy <- env$policyfn(model, reps = 10L)
```

@@

```{r}
# Simulate under the steady-state solution (given the model)
K = 1.5
alpha = 0.001
opt <- gym_conservation$models$fixed_action(env, fixed_action = alpha * 100 * 2 * K )
opt_sims <- env$simulate(opt, reps = 100L)
opt_policy <- env$policyfn(opt)
```

@@

```{r}
sims_df <- bind_rows(TD3_sims, opt_sims, .id = "model") %>%
  mutate(model = c("TD3", "steady-state")[as.integer(model)])

policy_df <- bind_rows(TD3_policy, opt_policy, .id = "model") %>%
  mutate(model = c("TD3", "steady-state")[as.integer(model)])

gamma <- 1 #discount
reward_df <- sims_df %>% 
  group_by(rep, model) %>%
  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
  group_by(time, model) %>%
  summarise(mean_reward = mean(cum_reward), 
            sd = sd(cum_reward), .groups = "drop")

@@

```{r}
# ensemble statistics
ymin <- function(x) last(x[(ntile(x, 20)==1)])
ymax <- function(x) last(x[(ntile(x, 20)==19)])

fig_sims <- 
  sims_df %>%
   group_by(time, model) %>% 
   summarise(ymin = ymin(state),
            ymax = ymax(state),
            state = mean(state), .groups = "drop") %>%
  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
   geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
```

```{r}
fig_policy <- 
  policy_df %>% ggplot(aes(state, action, 
                           group=interaction(rep, model),
                           col = model)) + 
  geom_line(show.legend = FALSE) 
```

@@

## First 4 of the TD3 reps falling below .2 for more than 15 steps
is_low <- sims_df %>% 
  filter(model == "TD3") %>% 
  group_by(rep, model) %>% 
  summarize(low = sum(state < .2) > 10) %>% 
  filter(low) %>% head(6)

@@

```{r}
# Some individual replicates, for comparison
is_low <- sims_df %>% filter(model == "steady-state") %>%
  group_by(rep, model) %>% summarize(low = sum(state < .2) > 10) %>% 
  filter(low) %>% head(6)
sims_df %>% inner_join(is_low) %>%
  ggplot(aes(time, state, group=interaction(model, rep))) +
  geom_line(color = pal[1], show.legend = FALSE) + facet_wrap(~rep)
```

>>>>

scale_colour_discrete = function(...) ggthemes::scale_colour_solarized()
scale_fill_discrete = function(...) ggthemes::scale_fill_solarized()
pal = ggthemes::solarized_pal()(8)
txtcolor = "#586e75"
```

# Overview

Development of any deep reinforcement learning application can be divided into two components: an _environment_ and an _agent_.
An environment is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
An agent is typically a computational neural network trained using an appropriate algorithm to suggest an action based on available information about the environment.

Somewhat confusingly, frameworks will often refer to a trained agent as a "model", while also referring to the approaches used to implement and train the agent as "model free" because the agent does not attempt to form an internal model of the environmental process.  We will attempt to mitigate this confusion by referring to the software component used to predict actions as an "agent", and reserving the use of "model" to refer to the ecological model(s) embedded in the environmental simulator.

@brockman2016 identified (1) the lack of standardization of environments used in publication and (2) the need for better benchmark environments as two core problems limiting more rapid advancement of RL research, and proposed the creation of the OpenAI `gym` framework to address both poblems.

A rapidly growing set of material for learning about deep reinforcement learning is now freely available.

@@

At the time of writing, several major frameworks exist which provide convenient access to baseline algorithms commonly used in deep reinforcement learning.
While we do not seek to provide a comprehensive review of available frameworks, some familiarity with current frameworks can be helpful.
Like many machine learning libraries, these frameworks are themselves each built around one of two popular machine learning libraries, PyTorch [@torch] or Tensorflow [@tensorflow], and all are based in Python.
Existing frameworks we evaluated include Keras-RL [@keras-rl], Tensorflow Agents [@tensorflow-agents],  OpenAI Spinning Up [@spinningup] OpenAI Baselines, Stable-Baselines2 [@sb2], and Stable-Baselines3 [@sb3]. 
Keras-RL saw widespread early adoption, but is built on Tensorflow version 1.x. It is incompatible with Tensorflow 2.x and not actively maintained.
Tensorflow Agents is developed by the Tensorflow team, a recent and actively developed framework with support for both Tensorflow 1.x and 2.x and good support for low-level customization of RL algorithms.
However, higher-level interfaces and high-level documentation are still relatively limited.
OpenAI's SpinningUp is an education-targeted framework useful for developers wanting to become more familiar with the internal methods of RL algorithms.  
OpenAI's Baselines is primarily Tensorflow 1.x-based implementation of many recently published RL algorithms.
Researchers at Ensta Paris Tech first created Stable Baselines as a fork of the OpenAI implementation, rigorously addressing numerous issues in documentation, testing, and coding style that have helped make their fork see even greater adoption. 
Stable-Baselines3 is the most recent version (Feb 2021), a ground-up rewrite which switches to a PyTorch-based implementation and further strengthens internal checks such as static types.
The examples here all use the Stable-Baselines3 framework, though researchers in this area should expect frameworks to continue to emerge and evolve. 
Grand challenge problems will likely require significant development beyond the current algorithms and capabilities available in existing frameworks.

# Building a simulation environment `gym`

[@openai_gym]

@@

## Python dependencies loaded via R
sb3         = import ("stable_baselines3")
gym         = import ("gym")
gym_fishing = import("gym_fishing")
gym_conservation = import("gym_conservation")
```

@@

```{r}
## reproducible settings
np = import("numpy")
seed = 24L # integer
np$random$seed(seed)
# Optionally set to "" to force CPU-evaluation if needing perfect reproducibility
Sys.setenv("CUDA_VISIBLE_DEVICES"="")
set.seed(seed)
```

@@

```{r }
## initialize the environment
env = gym$make("fishing-v1", sigma = 0.1)
```

@@

```{r}
a2c = sb3$A2C('MlpPolicy', env, verbose=0L, seed = seed) # L indicates a (Long) integer, not floating point
```

@@

```{r}
# Simulate management under the trained agent
a2c = sb3$A2C$load("cache/a2c")
```

@@

## represent the initial state size in the 'rescaled' state space.
state = env$get_state( 0.75 )
state
```

@@

  out = a2c$predict(state)
  action = out[[1]]
  result = env$step(action)
  state = result[[1]]

@@

```{r}
a2c_sims = env$simulate(a2c, reps = 500L)
```

@@

```{r}
a2c_policy = env$policyfn(a2c, reps = 50L)
```

@@

```{r}
# Simulate management under the trained agent
td3 = sb3$TD3$load("cache/td3")
td3_sims = env$simulate(td3, reps = 500L)
td3_policy = env$policyfn(td3, reps = 50L)
```

@@

```{r}
# Simulate under the optimal solution (given the model)
opt = gym_fishing$models$escapement(env)
opt_sims = env$simulate(opt, reps = 500L)
opt_policy = env$policyfn(opt)
```

@@

```{r}
sims_df = bind_rows(td3_sims, a2c_sims, opt_sims, .id = "model") %>%
  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])

policy_df = bind_rows(td3_policy, a2c_policy, opt_policy, .id = "model") %>%
  mutate(model = c("TD3", "A2C", "optimal")[as.integer(model)])

gamma = 1 #discount
reward_df = sims_df %>% 
  group_by(rep, model) %>%
  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
  group_by(time, model) %>%
  summarise(mean_reward = mean(cum_reward), 
            sd = sd(cum_reward), .groups = "drop")

@@

```{r}
ymin = function(x) last(x[(ntile(x, 20)==1)])
ymax = function(x) last(x[(ntile(x, 20)==19)])

fig_sims = 
sims_df %>% 
  group_by(time, model) %>% 
  summarise(ymin = ymin(state),
            ymax = ymax(state),
            state = mean(state), .groups = "drop") %>%
  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
  geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
```

```{r}
fig_policy = 
  policy_df %>% ggplot(aes(state, action, 
                           group=interaction(rep, model),
                           col = model)) + 
  geom_line(show.legend = FALSE) + 
  coord_cartesian(xlim = c(0, 1.3), ylim=c(0,0.9)) 
```

@@

fig_reward = reward_df %>% 
  ggplot(aes(time, mean_reward)) + 
  geom_ribbon(aes(ymin = mean_reward - 2*sd, 
                  ymax = mean_reward + 2*sd, fill = model),
              alpha=0.25, show.legend = FALSE) + 
  geom_line(aes(col = model), show.legend = FALSE) + 
  ylab("reward")
```

@@

```{r}
# Simulate management under the trained agent
env = gym$make("fishing-v1", r = 1.0379274,	K = 1.197693, sigma = 0.1121662)
td3 = sb3$TD3$load("cache/td3")

@@

```{r}
hake = read_csv("../data/hake.csv")
x0 = hake %>%
 filter(year == min(year)) %>%
  pull(scaled_biomass)
Tmax = 15
years =1986:2000

td3_state = td3_action = numeric(Tmax)
td3_state[1] = x0
td3_action[1] = NA

@@

```{r}
env$reset()
## represent the initial state size in the 'rescaled' state space.
state = env$get_state( x0 )

@@

  # RL-recommended harvest action:
  out = td3$predict(state)
  action = out[[1]]

  # Record state and resulting action
  td3_state[i] = env$get_fish_population(state)
  td3_action[i] = env$get_quota(action)

  # Implement RL-recommended harvest. 
  result = env$step(action)
  state = result[[1]]

  #state = env$get_state(hake$scaled_biomass[i+1])
}

df = bind_rows(
  tibble(year = years, state = td3_state, action = td3_action, model = "TD3"),
  tibble(year = years, state = hake$scaled_biomass, action = hake$scaled_catch, model = "historical"))

@@

```{r message=FALSE, fig.cap="Bifurcation diagram for tipping point scenario. The ecosystem begins in the desirable 'high' state under an evironmental parameter (e.g. global mean temperature, arbitrary units) of 0.19.  In the absence of conservation action, the environment worsens (e.g. rising mean temperature) as the parameter increases.  This results in only a slow degredation of the stable state, until the parameter crosses the tipping point threshold at about 0.215, where the upper stable branch is anihilated in a fold bifurcation and the system rapidly transitions to lower stable branch, around state of 0.1.   Recovery to the upper branch requires a much greater conservation investment, reducing the parameter all the way to 0.165 where the reverse bifurcation will carry it back to the upper stable branch."}
bifur_df = read_csv("figs/bifur.csv", col_types = "dccd")
bifur_df %>%
  ggplot(aes(parameter, state, lty=equilibrium, group = group)) + 
  geom_line()

@@

```{r}
env = gym$make("conservation-v6")
```

@@

OU = sb3$common$noise$OrnsteinUhlenbeckActionNoise
action_noise = OU(mean = np$zeros(1L),  sigma = noise_std * np$ones(1L))

@@

model = sb3$TD3$load("cache/td3-conservation")
TD3_sims = env$simulate(model, reps = 100L)
TD3_policy = env$policyfn(model, reps = 10L)
```

@@

```{r}
# Simulate under the steady-state solution (given the model)
K = 1.5
alpha = 0.001
opt = gym_conservation$models$fixed_action(env, fixed_action = alpha * 100 * 2 * K )
opt_sims = env$simulate(opt, reps = 100L)
opt_policy = env$policyfn(opt)
```

@@

```{r}
sims_df = bind_rows(TD3_sims, opt_sims, .id = "model") %>%
  mutate(model = c("TD3", "steady-state")[as.integer(model)])

policy_df = bind_rows(TD3_policy, opt_policy, .id = "model") %>%
  mutate(model = c("TD3", "steady-state")[as.integer(model)])

gamma = 1 #discount
reward_df = sims_df %>% 
  group_by(rep, model) %>%
  mutate(cum_reward = cumsum(reward * gamma^time)) %>%
  group_by(time, model) %>%
  summarise(mean_reward = mean(cum_reward), 
            sd = sd(cum_reward), .groups = "drop")

@@

```{r}
# ensemble statistics
ymin = function(x) last(x[(ntile(x, 20)==1)])
ymax = function(x) last(x[(ntile(x, 20)==19)])

fig_sims = 
  sims_df %>%
   group_by(time, model) %>% 
   summarise(ymin = ymin(state),
            ymax = ymax(state),
            state = mean(state), .groups = "drop") %>%
  ggplot(aes(time, state, ymin = ymin, ymax = ymax, fill=model)) + 
   geom_ribbon(alpha= 0.3) + geom_line(aes(col = model))
```

```{r}
fig_policy = 
  policy_df %>% ggplot(aes(state, action, 
                           group=interaction(rep, model),
                           col = model)) + 
  geom_line(show.legend = FALSE) 
```

@@

## First 4 of the TD3 reps falling below .2 for more than 15 steps
is_low = sims_df %>% 
  filter(model == "TD3") %>% 
  group_by(rep, model) %>% 
  summarize(low = sum(state < .2) > 10) %>% 
  filter(low) %>% head(6)

@@

```{r}
# Some individual replicates, for comparison
is_low = sims_df %>% filter(model == "steady-state") %>%
  group_by(rep, model) %>% summarize(low = sum(state < .2) > 10) %>% 
  filter(low) %>% head(6)
sims_df %>% inner_join(is_low) %>%
  ggplot(aes(time, state, group=interaction(model, rep))) +
  geom_line(color = pal[1], show.legend = FALSE) + facet_wrap(~rep)
```
